Parsing Postfix Log Files

### 2009-04-23 doc/logparser-acronyms.tex
Changes after running diction. Changed acknowledgements.

\newglossaryentry{DNSBL}{
    type=\acronymtype,
    name={DNSBL},
    description={DNS Blacklist},
    first={DNS Blacklist (DNSBL)}
}
\newglossaryentry{DNSBL glossary}{
    name={DNS Blacklist},
    description={
        A \acronym{DNSBL} is a simple collaborative anti-spam technique
        used to reject or penalise mail sent from IP addresses believed to
        be the source of large volumes of spam.  The criteria used when
        deciding if an IP address should be included vary widely between
        \acronym[s]{DNSBL}, so before using one it is very important to
        check their listing policies.  To use a \acronym{DNSBL}, Postfix
        makes a DNS request incorporating the IP address of the client; if
        the requested hostname is found the client is on the
        \acronym{DNSBL}.
    }
}

@@

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \glsreset{SMTP}\acronym{SMTP} is the protocol used for transferring
    mail between the sending and receiving \acronym{MTA}\@.  It is a
    simple, human readable, plain text protocol, making it quite simple to
    test and debug problems with it.  A detailed description of
    \acronym{SMTP} is beyond the scope of this thesis: the original
    protocol definition is in \acronym{RFC}~821~\cite{RFC821}, which was
    superceded by \acronym{RFC}~2821~\cite{RFC2821}.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

>>>>

\newglossaryentry{DNSBL}{
    type=\acronymtype,
    name={DNSBL},
    description={DNS Blacklist},
    first={DNS Blacklist (DNSBL)}
}
\newglossaryentry{DNSBL glossary}{
    name={DNS Blacklist},
    description={
        A \acronym{DNSBL} is a simple collaborative anti-spam technique
        used to reject or penalise mail sent from IP addresses believed to
        be the source of large volumes of spam.  The criteria used when
        deciding if an IP address should be included vary widely between
        \acronym[s]{DNSBL}, so before using one it is essential to check
        their listing policies.  To use a \acronym{DNSBL}, Postfix makes a
        DNS request incorporating the IP address of the client; if the
        requested hostname is found the client is on the \acronym{DNSBL}.
    }
}

@@

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \glsreset{SMTP}\acronym{SMTP} is the protocol used for transferring
    mail between the sending and receiving \acronym{MTA}\@.  It is a
    simple, human readable, plain text protocol, making it quite simple to
    test and debug problems with it.  A detailed description of
    \acronym{SMTP} is beyond the scope of this thesis: the original
    protocol definition is in \acronym{RFC}~821~\cite{RFC821}, later
    superceded by \acronym{RFC}~2821~\cite{RFC2821}.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

### 2009-04-23 doc/logparser-background.tex
Changes after running diction. Changed acknowledgements.

\SMTPglossaryDescription{} Despite the simplicity of the protocol, many
virus and spam sending programs fail to implement it properly, so requiring
strict adherence to the protocol specification is beneficial in protecting
against spam and viruses.\footnote{Originally all mail servers adhered to
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \acronym{RFC}~760~\cite{rfc760}, but unfortunately
that principle was written in a friendlier time.  Given the deluge of spam
that mail servers are subjected to daily, a more appropriate maxim could
be: \textit{Require strict adherence to relevant \acronyms{RFC}; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  It is neither as friendly nor as
catchy, but it more accurately reflects the current situation.} A typical
\acronym{SMTP} conversation resembles the following (the lines starting
with a three digit number are sent by the server, all other lines are sent
by the client):

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.  For example, a phishing mail
might claim that the payroll system had a disastrous disk failure, and
until the server is replaced all salary payments will have to be processed
manually, so please reply to this mail with your name, address, and bank
account details; the criminal can then use any details sent to him to help
with identity theft.

>>>>

\SMTPglossaryDescription{} Despite the simplicity of the protocol, many
virus and spam sending programs fail to implement it properly, so requiring
strict adherence to the protocol specification is beneficial in protecting
against spam and viruses.\footnote{Originally all mail servers adhered to
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \acronym{RFC}~760~\cite{rfc760}, but unfortunately
that principle was written in a friendlier time.  Given the deluge of spam
that mail servers are subjected to daily, a more appropriate maxim could
be: \textit{Require strict adherence to relevant \acronyms{RFC}; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  It is neither as friendly nor as
catchy, but it more accurately reflects the current circumstances.} A
typical \acronym{SMTP} conversation resembles the following (the lines
starting with a three digit number are sent by the server, all other lines
are sent by the client):

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.  For example, a phishing mail
might pretend that the payroll system had a disastrous disk failure, and
until the server is replaced all salary payments will have to be processed
manually, so please reply to this mail with your name, address, and bank
account details; the criminal can then use any details sent to him to help
with identity theft.

### 2009-04-23 doc/logparser-complications.tex
Changes after running diction. Changed acknowledgements.

Process scheduling can have a still more confusing effect: the child mail
will often be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  Unfortunately, it is not possible
to distinguish between child mails waiting to be tracked and other mails
whose origin is unknown (except for bounce notifications, as described in
\sectionref{identifying bounce notifications}); postponing the entry into
the database of mails whose origin is unknown can result in those mails
remaining indefinitely in the state tables, causing a queueid clash if the
queueid is reused, resource usage for the period they remain in the state
tables, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} can not be expected to fully understand and reconstruct a
mail when it is lacking some of that mail's log lines.

@@

        If all the checks pass the mail will be discarded from the state
        tables, and will not be entered in the database.  If one or more
        checks do not pass it is assumed that the timeout happened after a
        rejected delivery attempt.  This assumption is not necessarily
        correct, because Postfix could have accepted an earlier recipient,
        rejected a later one, and continued to accept delivery of the mail
        for the first recipient.  Tn that case the timeout applies to the
        partially accepted mail, which should be discarded; however, this
        has not occurred in practice.  Processing timeouts is further
        complicated by the presence of out of order \daemon{cleanup} log
        lines: see \sectionref{discarding cleanup log lines} for details.

@@

    \item Does the \texttt{timeout\_queueids} state table have an entry for
        the queueid in the log line?  If there is an entry, it will be
        removed, regardless of whether the remaining criteria are
        satisfied.

@@

    \item The timestamp of the \daemon{cleanup} log line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly, and
        empirical evidence shows that ten minutes is not unreasonable;
        hopefully it is a good compromise between false positives (log
        lines incorrectly discarded) and false negatives (new state table
        entries incorrectly created).

>>>>

Process scheduling can have a still more confusing effect: the child mail
will often be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is processed, and \parsername{} cannot warn the user;
instead they must be marked as ready for entry and subsequently entered
once their origin has been identified.  Unfortunately, it is not possible
to distinguish between child mails waiting to be tracked and other mails
whose origin is unknown (except for bounce notifications, as described in
\sectionref{identifying bounce notifications}); postponing the entry into
the database of mails whose origin is unknown can result in those mails
remaining indefinitely in the state tables, causing a queueid clash if the
queueid is reused, resource usage for the period they remain in the state
tables, and, most importantly, \parsernames{} understanding of such mails
is incorrect.  This problem has only been observed when a state table entry
is missing its initial log line, usually because it is in an earlier log
file; this specific instance is not a serious problem, because
\parsername{} cannot be expected to fully understand and reconstruct a mail
when it is lacking some of that mail's log lines.

@@

        If all the checks pass the mail will be discarded from the state
        tables, and will not be entered in the database.  If one or more
        checks do not pass it is assumed that the timeout happened after a
        rejected delivery attempt.  This assumption is not necessarily
        correct, because Postfix could have accepted an earlier recipient,
        rejected a later one, and continued to accept delivery of the mail
        for the first recipient.  In that case the timeout applies to the
        partially accepted mail, which should be discarded; however, this
        has not occurred in practice.  Processing timeouts is further
        complicated by the presence of out of order \daemon{cleanup} log
        lines: see \sectionref{discarding cleanup log lines} for details.

@@

    \item Does the \texttt{timeout\_queueids} state table have an entry for
        the queueid in the log line?  If an entry is found it will be
        removed, regardless of whether the remaining criteria are
        satisfied.

@@

    \item The timestamp of the \daemon{cleanup} log line must be within ten
        minutes of the mail acceptance timestamp.  Timeouts happen after
        five minutes, but some data may have been transferred slowly, and
        empirical evidence shows that ten minutes is not unreasonable; it
        appears to be a good compromise between false positives (log lines
        incorrectly discarded) and false negatives (new state table entries
        incorrectly created).

### 2009-04-23 doc/logparser-conclusion.tex
Changes after running diction. Changed acknowledgements.

A parser's ability to correctly parse its inputs is extremely important;
\parsernames{} coverage of \numberOFlogFILES{} log files, each containing
one day's log lines, is discussed in \sectionref{parsing coverage}.  Both
its success at recognising individual log lines  and its correctness in
reconstructing each mail's journey through Postfix are described in detail,
including the results of manually verifying that a randomly selected
portion of a log file was correctly parsed.  Experience implementing
\parsername{} shows that full input coverage is relatively easy to achieve
with this architecture, and that with enough time and effort a full
understanding of the input is possible.  Postfix log files would require
substantial time and effort to correctly parse regardless of the
architecture used; this architecture enables an iterative approach to be
used~\cite{stepwise-refinement}, as is practiced in many other software
engineering disciplines.

>>>>

A parser's ability to correctly parse its inputs is extremely important;
\parsernames{} coverage of \numberOFlogFILES{} log files, each containing
one day's log lines, is discussed in \sectionref{parsing coverage}.  Both
its success at recognising individual log lines  and its correctness in
reconstructing each mail's journey through Postfix are described in detail,
including the results of manually verifying that a randomly selected
portion of a log file was correctly parsed.  Experience implementing
\parsername{} shows that full input coverage is not difficult to achieve
with this architecture, and that with enough time and effort a full
understanding of the input is possible.  Postfix log files would require
substantial time and effort to correctly parse regardless of the
architecture used; this architecture enables an iterative approach to be
used~\cite{stepwise-refinement}, as is practiced in many other software
engineering disciplines.

### 2009-04-23 doc/logparser-implementation.tex
Changes made based on Carl's comments.
Changes after running diction. Changed acknowledgements.

\parsername{} deals with all the eccentricities and oddities of parsing
Postfix log files, presenting the resulting data in a normalised, simple to
use representation.  Unfortunately, dealing with the complications that
arise sometimes requires the parser to discard log lines (e.g.\
\sectionref{aborted delivery attempts}), and, less frequently, to discard a
data structure (e.g.\ \sectionref{timeouts during data phase}).
\parsername{} can parse log files from Postfix version 2.2 through to
version 2.5, and should be capable of parsing log files from later versions
with only minor modifications or an updated ruleset.  The parser's task is
to follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.  The intermingling of
log lines from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must handle multiple mails in
parallel, each potentially at a different stage in its journey, without any
interference between mails --- except in the minority of cases where
intra-mail interference is required, e.g.\ mail re-injected for forwarding
(\sectionref{Re-injected mails}).  The best way to deal with intermingling
of log lines is to maintain state information for every unfinished mail,
and to manipulate the appropriate mail correctly for each log line
encountered.

@@

\parsername{} makes a small number of (hopefully safe and reasonable)
assumptions:

@@

@@

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible; one of the few limitations is that \parsername{} is
written in Perl, so new actions must also be written in Perl.  The
implementer writes a subroutine that accepts the standard arguments given
to actions, and registers it with the framework by calling the framework's
\texttt{add\_actions()} subroutine.  The new action must be registered
before the rules are loaded, because it is an error for a rule to specify
an unregistered action; this helps catch mistakes made when adding new
rules.  No other work is required from the implementer to integrate the
action into the parser; all of their attention and effort can be focused on
correctly implementing their new action.  The action may need to extend the
list of valid combinations described in \sectionref{out of order log lines}
if the new action creates a different set of acceptable programs, but this
would only be necessary if the new action processes log lines from Postfix
components that \parsername{} does not have rules for, e.g.\
\daemon{virtual} or \daemon{lmtp}.

>>>>

\parsername{} deals with all the eccentricities and oddities of parsing
Postfix log files, presenting the resulting data in a normalised, simple to
use representation.  Unfortunately, dealing with the complications that
arise sometimes requires the parser to discard log lines (e.g.\
\sectionref{aborted delivery attempts}), and, less frequently, to discard a
data structure (e.g.\ \sectionref{timeouts during data phase}).
\parsername{} can parse log files from Postfix version 2.2 through to
version 2.5, and should be able to parse log files from later versions with
only minor modifications or an updated ruleset.  The parser's task is to
follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.  The intermingling of
log lines from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must handle multiple mails in
parallel, each potentially at a different stage in its journey, without any
interference between mails --- except in the minority of cases where
intra-mail interference is required, e.g.\ mail re-injected for forwarding
(\sectionref{Re-injected mails}).  The best way to deal with intermingling
of log lines is to maintain state information for every unfinished mail,
and to manipulate the appropriate mail correctly for each log line
encountered.

@@

\parsername{} makes a few (hopefully safe and reasonable) assumptions:

@@

\label{framework in implementation}

@@

Adding new actions is not as facile as adding new rules, though care has
been taken in the architecture and implementation to make adding new
actions as painless as possible; one of the few limitations is that
\parsername{} is written in Perl, so new actions must also be written in
Perl.  The implementer writes a subroutine that accepts the standard
arguments given to actions, and registers it with the framework by calling
the framework's \texttt{add\_actions()} subroutine.  The new action must be
registered before the rules are loaded, because it is an error for a rule
to specify an unregistered action; this helps catch mistakes made when
adding new rules.  No other work is required from the implementer to
integrate the action into the parser; all of their attention and effort can
be focused on correctly implementing their new action.  The action may need
to extend the list of valid combinations described in \sectionref{out of
order log lines} if the new action creates a different set of acceptable
programs, but this would only be necessary if the new action processes log
lines from Postfix components that \parsername{} does not have rules for,
e.g.\ \daemon{virtual} or \daemon{lmtp}.

### 2009-04-23 doc/logparser.tex
Corrections after proofreading chapters 6 & 7 and the appendices.
Changes after running diction. Changed acknowledgements.

\noindent{}I would like to express my appreciation to my supervisor, Dr.\
Carl Vogel, for his advice, assistance, and guidance.

@@

\noindent{}I would like to thank my wife, Ariane Tobin, for her patience,
good will, and encouragement.

@@

% Add some glossary entries that should be present, but lack an appropriate
% place in the text to mark them.
\glsadd{queueid}
\renewcommand{\glossarypostamble}{\label{Glossary}}
\printglossary[style=nospacelist]{}
\renewcommand{\glossarypostamble}{\label{Acronyms}}
\printglossary[type=\acronymtype,style=eqlist]{}
\renewcommand{\glossarypostamble}{\label{Postfix Daemons}}
\printglossary[type=postfix,style=nospacelist]{}

>>>>

@@

\noindent{}I am indebted to my supervisor, Dr.\ Carl Vogel, for his advice,
assistance, and guidance.

\noindent{}I am grateful to my wife, Ariane Tobin, for her patience, good
will, and encouragement.

@@

% Add some glossary entries that should be present, but lack an appropriate
% place in the text to mark them.
\glsadd{queueid}
\glsresetall{}
\renewcommand{\glossarypostamble}{\label{Glossary}}
\printglossary[style=nospacelist]{}
\renewcommand{\glossarypostamble}{\label{Acronyms}}
\printglossary[type=\acronymtype,style=eqlist]{}
\glsresetall{}
\renewcommand{\glossarypostamble}{\label{Postfix Daemons}}
\printglossary[type=postfix,style=nospacelist]{}
